---
title: "NYPD Shooting Report Data Analysis"
author: "Anya V"
date: "07/24/2024"
output: 
   html_document: default
   pdf_document: default
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(lubridate)
library(ggplot2)
```

## R Markdown

This is a markdown for NYPD Shooting Incident Data(Historic)

First, we will download the dataset from the government catalog at https://catalog.data.gov/dataset

```{r nypd_shooting_data}
url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?"

# Read and print raw data from the csv
shootings_raw <- read_csv(url_in)
summary(shootings_raw)

# Cleanup columns. We will focus on shootings occurrences by date, location, age, sex, and race of victims and perpetrators
shootings <- shootings_raw %>% mutate( DATE = mdy(OCCUR_DATE)) %>% rename( TIME = OCCUR_TIME,PERP_AGE = PERP_AGE_GROUP, VIC_AGE = VIC_AGE_GROUP) %>% select ( DATE, TIME, BORO, PERP_AGE, PERP_SEX, PERP_RACE, VIC_AGE, VIC_SEX, VIC_RACE)
print( shootings, n = 25)

# Examine missing data in race, age and gender columns. As we can see in the tables substantial amount of data is missing
# from all three categories, making analysis and modeling impossible for these criterias.
shootings_race <- shootings %>% select( PERP_RACE) %>% group_by( PERP_RACE) %>% summarize( cases = n())
print( shootings_race)
shootings_age <- shootings %>% select( PERP_AGE) %>% group_by( PERP_AGE) %>% summarize( cases = n())
print(shootings_age)
shootings_gender <- shootings %>% select( PERP_SEX) %>% group_by( PERP_SEX) %>% summarize( cases = n())
print( shootings_gender)


# Im curious if there are seasonal variations in the number of shootings
# Analysis: we can see from the graph that there are significantly, two to three times, more cases in the summer months. 
# This data is consistent over the 18 years period. 
shootings_by_month <- shootings %>% mutate( year = year(DATE), month = month(DATE)) %>% group_by( year,month) %>% summarize( cases = n())
# print( shootings_by_month)
# Graph a month to month number of cases for each year
shootings_by_month %>% ggplot(aes(x = month, y = cases, color = as.factor(year), group = year)) + 
  geom_line() +
  labs( title = "Number of Shootings by Month", x = "Month", y = "Number of Cases", color = "Year") +
  scale_x_continuous( breaks = 1:12, labels = month.abb)

# Now lets see if there is a trend over the last 18 years 
# Analysis: we can see that from 2006 to 2019 there was consistent and significant decline in the number of shootings per year.
# However, in 2020 and 2021 the number of cases jumped back to the 2006 levels. It is interesting that these years coincide
# with covid pandemic and shelter in place orders during these years. 
# We can observe the trend in the last two years the number of shootings plunging down back to pre-pandemic levels.
shootings_by_year <- shootings %>% mutate( year = year(DATE)) %>% select(year) %>% group_by(year) %>% summarise( cases = n())
#print(shootings_by_year)

shootings_by_year %>% ggplot(aes( x= factor(year), y = cases)) +
   geom_bar( stat = "identity", fill = "dark red") +
   labs( title = "Number of Shootings by Year", x = "Year", y = "Number of Cases") 

# Modeling: now lets create a model to show year to year tendencies in number of shootings
# There have been a general decline in the number of shootings year after year, however,
# during the covid lock down, the number of cases climed back up for two years. 

mod <- lm( cases ~ year, data = shootings_by_year )
shootings_by_year <- shootings_by_year %>% mutate( pred = predict(mod))
shootings_by_year
shootings_by_year %>% ggplot(aes( x = factor(year))) + geom_point( aes( y = cases), color = "blue") +
  geom_point( aes( y = pred), color = "red") +
  labs( title = "Number of Shootings Prediction Model", x = "Year", y = "Cases")


```

## Conclusion

The data set on New York shootings is covering a period from 2006 through 2023. 
There are several ways this data can be analyzed, such as by burrow, age, race 
and gender of the perpetrators, seasonal and yearly tendencies. 

We observed that significant, nearly half of shootings data has missing values
in age, race and gender columns. While we don't know the cause for incomplete 
data, it could be related to biases of persons recording the incidents. To 
mitigate the bias, additional data is needed. 

We chose to observe seasonal and year-to-year tendencies for which the data is
complete. As we saw in the charts, there is a strong seasonal correlation 
between the number of cases and seasons, with summer months having the peak of 
shootings, and winter the least. 

We observed that year to year tendency is the decline of number of shootings 
with exception for two pandemic years when the effects of pandemic and social 
isolation may have played a big role in uptake of cases.

I have not identified any personal biases, but it doesn't mean that there aren't
any. Usually people are blind to their biases. For the sake of this assignment,
lets assume that Im biased about covid-19 lock down policies and it's impact 
on mental health and consequently, violence, including homicide.
 
To mitigate this bias, I reviewed some data and read several research articles 
from NCBI( National Center for Biotechnology Information) and FBI websites.

I found out that while lock down measures were well-intentioned, one unintended 
adverse consequence was an increase in violence, including gun violence.
Social isolation on top of the stress of the pandemic had in impact on mental 
health is a cause of the rise in violent crime in 2020-2021 on the magnitude of 
25% increase over the previous years. Especially large cities were affected.
This finding alignes well with our model for NY shootings.


